PhD Proposal

### 2022-05-08 abstract/abstract.tex
continuo abstract

Thanks to the last developments in sequencing technologies, which had led both

@@

million variants between those human genomes. Among these variants 84.7 million

@@

\textit{pangenome} looking for associations between genetic variants, in order,
for example, to study also deseases outocomes.
In this particular historical period it is also impossible not to mention the
viruses. In fact, by nature, during infections, viruses replicate a lot but
often in a not perfect way, thus producing many inexact clones, referred as
\textbf{viral haplotypes} which, taken together, form the \textbf{viral
  pangenome}. The identification of all this haplotypes is crucial both for the
effective study of the spread of the virus and for the production of effective
drugs, in a content of high pharmacological resistance. \\

@@

A very intriguing tool, talking about \textit{GWAS}, to be cited is

@@

of \textit{PBWT} proved inevitable, further demonstrating the importance of this
data structure.\\

@@

proposed a use of his data structure in \textit{GWAS} context. Infact they

@@

of natural selection to build the so-called \textbf{ancestral recombination
  graph}, that contains complete informations about samples history.\\

@@

developing a simple new data structure.\\  

@@

Especially during my master’s degree I have deepened both the most theoretical
issues related to algorithms, especially in the field of bioinformatics, that
the most modeling and inference issues, related to the field of computational
systems biology, so it is my interest to continue my studies with the PhD

@@

context. Summarizing, some of the research goals of my \textit{Phd} are:

@@

  \item \textbf{Multiallelic data}. Thanks to the increasing growth in the
  production of genotype data the number of multiallelic sites is expected to
  grow, as the number of samples increases, even in the human
  genome, although now, for example, there is only a 2\% presence of triallelic
  sites. Moreover, not only could such sites be more than expected but they are
  usually not considered by most tools. A first step in this direction was made
  by Naseri with the already cited \textit{m-PBWT}. From the point of view of
  spatial complexity, especially about stored arrays for the \textit{FM-index},
  a run-length encoded version of the same could 
  allow the management of increasingly large data for the imputation phases. 
  In this context the aim is to implement a new 
  version of the \textit{RLPBWT} that can handle multiallelic data, adapting the
  current use of bitvectors to handle efficiently the \textit{LF-mapping} also
  with more than two alleles. 

@@

  handle, is a \textit{RLPBWT} version that admits the presence of

@@

  gaps, both mainly due to sequencing machines that are not yet perfect,
  although they now have very high correctness rates. Unfortunately, although 
  most of them are corrected as a preliminary step (mostly by heuristic
  methods), this is still an open problem. The development of a data structure
  based on the \textit{PBWT}, and as a next step on the \textit{RLPBWT}, that
  can manage missing data would allow the 
  creation of more and more complete tools for the study of haplotypes and for
  the various \textit{GWAS}. 
  Handling missing data is known to be \textit{hard} so I should
  probably deal with parametric algorithms or approximate algorithms, based on

@@

  \item \textbf{Genotype imputation}. As introduced, having developed a
  wide-range interest in computer science potential in biological studies, I do

@@

  and develop during my PhD. Including the management of missing data and
  multiallelic data can further improve the state of the art of genotype
  imputation, allowing even more precise inference of unobserved genotypes, and,
  in more detail, of \textit{GWAS}. All this must be read in the perspective of
  a continuous increase of the available data, moreover not not only regarding
  the human sequencing.  

>>>>

Thanks to the last developments in sequencing technologies, with \textbf{Next
  Generation Sequencing (\textit{NGS})} and \textbf{third-generation
  sequencing}, which had led both 

@@

million variants between those human genomes. Among these variants, 84.7 million

@@

\textit{pangenome} looking for associations between genetic variants, for
example in order to study outocomes of diseases.
In this particular historical period, it is impossible not to mention
viruses. In fact, by nature, viruses replicate a lot but
often in a not perfect way, during infections. Thus produces many inexact
clones, referred as 
\textbf{viral haplotypes}, which, taken together, form the \textbf{viral
  pangenome}. The identification of all these haplotypes is crucial both to the
study of the spread of viruses and to the production of efficient
drugs, in a context of high pharmacological resistance. \\

@@

A very intriguing tool, talking about \textit{GWAS}, is

@@

of \textit{PBWT} has been proved to be inevitable, further demonstrating the
importance of this data structure.\\

@@

proposed a use of his data structure in \textit{GWAS} context. In fact, they

@@

of natural selection and to build the so-called \textbf{ancestral recombination
  graph}, that contains complete information about history of samples.\\

@@

developing a simple new data structure.

@@

During my master’s degree, I have deepened some theoretical
issues related to algorithms, especially in bioinformatics, and
some modeling and inference topics, related to computational 
systems biology. So, it is my interest to continue my studies with the PhD

@@

context.\\
Summarizing, the most important research goals of my PhD are:

@@

  \item \textbf{Multiallelic data}. Thanks to the growth in the
  production of genotypic data, the number of multiallelic sites is expected to
  grow, as well as the number of sample (even if there is only a 2\% presence of
  triallelic sites, actually known in the human genome). Moreover, not only
  such sites could be more than expected but they are 
  usually not considered by the majority of tools. A first step in this
  direction was made 
  by Naseri with the already cited \textit{m-PBWT}.
  Talking about spatial complexity, run-length encoded version of the stored
  arrays for the \textit{FM-index} could allow the management of increasingly
  large data for the imputation phases.   
  In this context, the aim is to implement a new 
  version of the \textit{RLPBWT} that can manage multiallelic data, adapting the
  current use of bitvectors to handle efficiently the \textit{LF-mapping},
  with more than two alleles, too. 

@@

  design, is a \textit{RLPBWT} version that admits the presence of

@@

  gaps, both mainly due to the imperfections of sequencing technologies,
  although now they have really high correctness rates. Unfortunately, this is
  still an open problem, even if most of the errors are corrected in a
  preliminary step (mostly by heuristic methods).
  Handling missing data is known to be \textit{hard} so, I should
  probably deal with parametric or approximate algorithms, based on

@@

  \item \textbf{Genotype imputation \& other goals}. As introduced, I have
  developed a wide-range interest in the application of computer science to
  biology. So, I do 

@@

  and develop during my PhD. The management of missing and
  multiallelic data can further improve the state of the art of \textit{GWAS}
  and of genotype imputation, allowing even more precise inference of unobserved
  genotypes. All this must be read in the perspective of 
  a continuous rise of available data, not only regarding the human
  sequencing. \\
  I would also  point out my interest in deepening the issues related to
  pattern matching, indexing structures and pangenome graph. During
  my PhD, it’s my intention to focus on all the BIAS laboratory's topics of
  interest, from the point of view of bioinformatics and experimental 
  algorithmics. 
